---
title: "Variable Storage & Data Structures"
description: | 
  Getting to know the building blocks of data science in R
output: 
  distill::distill_article:
    toc: true
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = FALSE)
library(knitr)

```

:::obj

**Learning objectives**
 
 - Assigning objects
 - Know basic object classes (logical, numeric, factor, character)  
 - Understand basic R data structures (vectors, lists, data.frames)
 - Know properties of R data structures and how to interact with them
 - Know how to spot missing values (`NA`)

:::

## Assignment

R is an object oriented programming language, which means that it is oriented around objects which can be data (e.g., data.frames, vectors, lists) or code (e.g., functions). We've actually been using assignment already to create variables, or objects. In the previous lesson, we read in csv files and assigned them to a variable using the assignment operator `<-` (shortcut: `Alt` + `-`). Avoid using `=`, which will also work but cause confusion later ([here's a blog post on the history behind `<-` for later](https://colinfay.me/r-assignment/)). 

When assigning an object, avoid overly simplistic names (e.g., `x`, `y`), and rather, pick concise names that describes the object and improve code interpretability.  

```{r assignment}
library(readr)
library(here)

# read the stations.csv file and assign it to an object "stations"
stations <- read_csv(here("data", "gwl", "stations.csv"))

# function that reports the class of each column of an input dataframe
get_column_classes <- function(df){
  col_classes <- sapply(df, class)
  return(col_classes)
}

# run the function to see the output
get_column_classes(stations)
```

Don't worry about how we created a function here -- there's an entire [module on functions]() coming up. We only define a custom function to illustrate that code too can be a object. Notice that in the Global Environment pane, we have "Data" including the object `stations` and "Functions" containing the custom function `get_column_classes()`.


## Basic object classes

Every object in R has a `class` property, and each property defines what functions will work on it. Understanding how to check the class of an object and figure out what functions can be applied to what object classes is paramount. Many bugs are the result of functions applied to the wrong object class.

There are 4 major object classes, arranged in order of complexity:

* logical (`TRUE`, `FALSE`)  
* numeric  
    + integer (e.g., 0, 1, 2, ...)  
    + double: AKA float64, a computer number format usually occupying 64 bits in memory  
* factor (categorical and ordinal variables)  
* character (strings)  

### logical


### numeric


### factor


### character




## Basic data structures


### vector

property length

### list

property length


### data.frame and tibble

property length refers to number of cols
property dim, nrow, ncol




## A note on `NA`

`NA` has a special meaning in R and designates a missing value. 

```{r}
et <- c(2, 2, NA, 4)
mean(et)
median(et)
sum(et)
```

