---
title: "Getting State & County Boundaries"
output: 
  bookdown::html_document2:
    code_fold: show
    includes:
      in_header: assets/header.html
      after_body: assets/footer.html
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(knitr)

```

<style>
  .title{
    display: none;
  }
</style>

<br>
<br>

:::obj

 **Objectives:**
 
 - Use {`USAboundaries`} package to download and save boundary data for State and Counties
 - save out data in different spatial formats: (`.shp`, `.geojson`, `.kml`, `.csv`)
 
:::

<br>

## Downloading US Boundaries

There are several options to do this, and many work equally well. We'll demonstrate one possible option here, the `USAboundaries` package.

<br>

### **Loading Spatial Packages**

Make sure you load the following packages

```{r loadsf, echo=T, eval=T}

# load packages or "libraries"
library(tidyverse) # wrangling/plotting tools
library(sf) # "simple features" spatial package for vector based data
library(USAboundaries) # data for USA boundaries

```


<br>

### **Get State Boundary Data**

Now lets grab some data using a package called `USAboundaries`, which allows us to quickly download county or state outlines, in `sf` formats. Very handy for making some quick professional maps. 

First we'll grab a State boundary for CA, OR, and NV so we can filter these down later.

```{r getCA, echo=T, eval=T}

library(USAboundaries)

# Pick a State
state_names <- c("california", "oregon", "nevada") # a vector of names

# Download STATE data and add projection
states <- us_states(resolution = "high", states = state_names) # what does resolution do?

st_crs(states) # double check the CRS

# make a quick plot!
plot(states$geometry)

```

<br>

### **Save as `.shp`**

A commonly used geospatial file format is the **shapefile**. Unfortunately this file type usually contains 4-5 or more associated files for each spatial object. So, when saving out as `.shp`, expect to see at least 4 files with the same name but different file extensions. All will be needed to spatially orient and read the data back in.


```{r toSHP, echo=T, eval=T}

# save out as "shp"
st_write(states, "data/states_boundaries.shp",
         delete_dsn = TRUE) # setting to TRUE allows overwrite

```

<br>

### **Save as `.kml`**

`kml` or *Keyhole Markup Language* is an XML format which is used by Google Earth and other programs. It can contain multiple layers (similar to `.gpx`), and retains the CRS associated with the data. Here we are writing a single layer (our states). But we could easily add (or `append=TRUE`) additional data like points or lines to this same file.

```{r toKML, echo=T, eval=T}

# save out as "kml"
st_write(states, "data/states_boundaries.kml", 
         delete_dsn = TRUE) # setting to TRUE allows overwrite

# to read it back in...
states <- st_read("data/states_boundaries.kml")

```

<br>

### **Get County Boundaries**

Let's get counties for all of California, and then filter to a single county. Notice we give the `us_counties` function only one state, but could easily pass a vector of states as we did in the section above. 

```{r getCAcounties, echo=T, eval=T}

# Download outlines of all the counties for California
ca_counties<-us_counties(resolution = "high", states = "CA") 

# Pick a county of interest
#co_names <- c("El Dorado")  

# filter to just that county:
#eldor_co <- filter(ca_counties, name==co_names)

# save out
#st_write(eldor_co, "data/eldorado_county_boundary.shp")
```

<br>

### **Save as `.geojson`**

An open source file type that requires a single file, is compatible with vector data, and is based on JSON. There are other packages that can be used to read/write these files, but we'll stick with `sf` here.

```{r toGEOJSON, echo=T, eval=T}

st_write(ca_counties, "data/ca_counties_boundaries.geojson", 
         delete_dsn = TRUE)

```


### **Save as `.csv`**

We can always fall back on writing to a comma-delimited file, but with the inclusion of a `geometry` column as WKT (Well-known-text). This means the spatial data is written into the `.csv`, and then will need to be read in as such. Note, this format has a smaller file size compared with `geojson` and is comparable to `.kml`.

One downside is typically the CRS information isn't retained in the csv, so you need to know what CRS the data has and then set the CRS using `st_set_crs`

```{r toWKT, eval=T, echo=T}

# This writes the geometry column as a WKT (all in one column)
st_write(ca_counties, "data/ca_counties_boundaries.csv", 
         layer_options = "GEOMETRY=AS_WKT",
         delete_dsn = TRUE)

# to read in:
ca_counties2 <- st_read("data/ca_counties_boundaries.csv", 
                        stringsAsFactors = FALSE, 
                        as_tibble = TRUE,
                        options="GEOM_POSSIBLE_NAMES=WKT") %>% 
  # drop the WKT column
  dplyr::select(-WKT)

# also need to set the CRS:
ca_counties2 <- st_set_crs(ca_counties2, 4326)

```

<br>

Make sure this all worked by previewing with `tmap`.

```{r tmap, eval=T, message=FALSE, warning=FALSE}

library(tmap)
tmap_mode(mode = "view")

tmap::tm_shape(ca_counties2) +
  tm_polygons(alpha=0.1, col="seagreen") +
  tm_shape(states) + 
  tm_polygons(lwd=2, border.col = "gray10", alpha = 0, col=NA)
```

