---
title: "Download NHD Streamlines"
output: 
  bookdown::html_document2:
    code_fold: show
    includes:
      in_header: assets/header.html
      after_body: assets/footer.html
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(knitr)

```

<style>
  .title{
    display: none;
  }
</style>

<br>
<br>

:::obj

 **Objectives:**
 
 - Use {`nhdplusTools`} package to download and save basin and streamlines based on a location
 - Save out data into a `.geopackage`
 
:::

<br>

## Working with Watershed Data

For folks who work in rivers and streams, an immensely useful R package for pulling hydrology data from NHD is [`{nhdplusTools}`](https://usgs-r.github.io/nhdplusTools/). This package provides access to NHD flowlines, attributes, and basin characteristics, all within an `{sf}` framework.

Let's use this to grab some flowlines (riverlines), identify USGS gage stations within a basin, and save these out for a future analysis.

<br>

### **Loading Spatial Packages**

Make sure you load the following packages.

```{r loadsf, echo=T, eval=T}

# load packages or "libraries"
library(tidyverse) # wrangling/plotting tools
library(sf) # "simple features" spatial package for vector based data
library(nhdplusTools) # data from NHD
library(mapview)
library(tmap)
```


<br>

### **Download Basins based on USGS Gage**

Here we can use the **[{`nhdplusTools`}](https://usgs-r.github.io/nhdplusTools/)** package to grab watershed basin boundaries based on any USGS gage. You can find a map of USGS gages [here](https://maps.waterdata.usgs.gov/mapper/index.html).

For this example, we'll use a gage on the South Fork American River, near Placerville, CA. This is Gage 11444500. Let's break this into a few steps.

  1. The `nhdplusTools` package needs a `list` that identifies the type of data we're looking for (here a `nwissite`). This could also be a latitude and longitude starting point. For USGS sites, we also need to provide the ID for that site.
  2. Once we have this information, we can do a few things. To download the basin upstream of a point or gage we can use the `get_nldi_basin` function.


```{r getBasin, echo=T}

library(nhdplusTools)

# first identify the gage of interest: This on on the American
nldi_nwis <- list(featureSource = "nwissite", featureID = "USGS-11444500")

# next download the basin:
basin <- get_nldi_basin(nldi_feature = nldi_nwis)

```

  3. The NHD dataset contains an immense amount of information, but it is built on various ID's that can be linked. Every single stream segment in the US has a unique `COMID` which can be used. Let's find the `COMID` associated with the `USGS-11444500`. Note, the CSCI data has a COMID column which makes it possible to download data via the COMID for each site.
  
```{r getCOMID}

# find out comid of the gage or point:
(comid_pt <- discover_nhdplus_id(nldi_feature = nldi_nwis))

```

### **Get Streamline Data**

What's really great is we can use a few different options to download NHD streamline info. Using either a gage ID, a COMID, or even a lat-long!

We also can specify whether we want upstream or downstream mainstem or tributaries. `UT`= "Upstream Tributaries", `UM`="Upstream Mainstem", `DM`="Downstream Main", or `DD`="Downstream Diversions". We can also specify a distance we want to travel from that initial starting point. The default is 10km.

 > Here we pull downstream only from a CSCI COMID.

```{r getDM}

# use the comid to download streamlines U/S or D/S of the gage
# here downstream mainstem segments 50 km from the starting COMID
ds_main <- navigate_nldi(list("comid", "14983602"),
                                mode="DM",
                                distance_km =  50)

# let's see what downstream diversions look like:
ds_div <- navigate_nldi(list("comid", "14983602"),
                                mode="DD",
                                distance_km =  50)

```

```{r quickMap, echo=T, eval=F}

mapview(ds_div, legend=FALSE) + mapview(ds_main, legend=FALSE)

```

```{r tmap1Hide, eval=T, echo=F, message=FALSE, warning=FALSE}

tmap_mode(mode = "view")

tmap::tm_shape(ds_div, name = "DS Diversions") +
  tm_lines(col="darkblue", lwd = 2, legend.show = T) +
  tm_shape(ds_main, name="DS Mainstem") +
  tm_lines(col="skyblue", lwd=1, legend.show = T)

```



 > Here we pull upstream mainstem streamlines from a USGS Gage and a X/Y COMID.

```{r getUM, eval=F}

# pull mainstems from a USGS gage
us_main <- navigate_nldi(nldi_feature = nldi_nwis,
                                mode="UM",
                                distance_km =  100)

# pull mainstems from an X/Y (lat/lon) point, needs to be st_sfc
xypoint <- st_sfc(st_point(c(-120.616, 38.81448)), crs = 4326)
xycomid <- discover_nhdplus_id(xypoint)

us_tribs <- navigate_nldi(
  nldi_feature = list(featureSource="comid", featureID=xycomid),
  mode="UT",
  distance_km = 20)


```

### **Find Additional Gages**

Another great feature of `nhdplusTools` is the ability to search within the NHD network for additional NWIS sites. Let's search upstream of USGS-11444500 for additional gages.

```{r usGages}

# find other upstream gages or NWIS sites
us_gages <- navigate_nldi(nldi_feature = nldi_nwis, 
                          mode="UM",
                          data_source = "nwissite",
                          distance=100)

```

We can plot this all together!

```{r plot, eval=F}

plot(basin$geometry, border = alpha("darkblue", alpha = 0.5), lwd=2)
plot(ds_main$geometry, col="darkblue", lwd=3, add=T)
plot(ds_div$geometry, col="skyblue", add=T)
plot(us_main$geometry, col="dodgerblue", add=T)
#plot(us_tribs$geometry, col="purple", add=T)
plot(us_gages$geometry, bg="maroon", pch=21, add=T)

```

<br>

```{r tmap, eval=F, echo=F, message=FALSE, warning=FALSE}

library(tmap)
tmap_mode(mode = "view")

tmap::tm_shape(ca_counties2) +
  tm_polygons(alpha=0.1, col="seagreen") +
  tm_shape(states) + 
  tm_polygons(lwd=2, border.col = "gray10", alpha = 0, col=NA)
```

