---
title: "4. Importing and Exporting Data"
description: |
  A journey of ten thousand steps begins with a single <font color="#009E73">**import**</font> and ends with an <font color="#0072B2">**export**</font>.
output: 
  distill::distill_article:
    toc: true
creative_commons: CC BY
preview: https://r4wrds.com/img/cover.png
---



:::obj

**Learning objectives**
 
 - Know the basics of how a function works
 - Understand how to read (import) data into R
 - Explore data within R using `View()`
 - Understand how to write (export) data from R
 - Understand how to read and write different data formats (e.g., excel, csv, shp, rds, dbf, access)
 
:::

<br>

## What's a function?

Functions are how work gets done in `R`, and before we jump into reading and writing data, we need to know how functions work because we will use functions to perform these tasks.

A **function** takes any number of **arguments** and, performs some transformations, and returns an **output**. 

For example, the `R` function `sum()` takes any number of numeric arguments and adds them together (recall you can view the documentation for `sum()` by entering `?sum`). Let's add 1 and 2 like so:  

```{r function-example-sum}

sum(1, 2)

```

In `R`, we can create sequences easily. If we wanted to create a sequence of numbers from 1 to 10, we can use the function `seq()`, which takes 3 arguments: `from` (start value), `to` (end value), and `by` (increment of the sequence). 

```{r sequence}

seq(0, 10, 1)

```

Convince yourself that creating sequences of arbitrary length is possible. Can you create a sequence from 0 to 1 million by an increment of 500?

Because creating sequences incremented by 1 are so common, there's a special shorthand for these sequences, `1:10`. Let's take the sum of the sequence from 1 to 10 by providing it as an argument to the function `sum()`:

```{r sequence-sum}

sum(1:10)

```


`sum()` and `seq()` are two of many functions you'll encounter in `R`. Like all functions, they take inputs (arguments) and return an output. To take advantage of functions, we need to apply them to our data. Let's now use import and export functions in R to explore some water resources data.  

<br>

## Read (import) data

Data come in many formats, and `R` has utilities for reading and writing all kinds of data. In this lesson, we'll explore some of the most common data formats you'll encounter in the wild, and the functions used to import these data.

### csv (comma separated values) 

The comma separated value, or `.csv`, is a simple and effective way to store tabular data. To read a csv file, we first import the {`readr`} library, which contains the function `read_csv()`. We also load {`here`} in order to easily navigate our working directory. Let's read a file from our `data/gwl` folder that contains station information for groundwater level monitoring sites in California. We use the "relative path" created by `here()`, which encode the much longer "absolute path". You can also type "data/" and press **Tab** with your cursor just after the "/" to view all files in that path.

```{r read-csv}
library(here)
library(readr)

# read the "stations" csv, save it as an object called "stations", and print the object
stations <- read_csv(here("data", "gwl" ,"stations.csv"))

head(stations)

```

You can also pass a URL to `read_csv()`.

```{r read-github}
# read the "stations" csv from the Github URL
stations <- read_csv("https://raw.githubusercontent.com/r4wrds/r4wrds-data/main/intro/data/gwl/stations.csv")
```


R tells us upon import that this data has 43,807 rows and 15 columns. We can verify this with the `nrow()` and `ncol()` functions, and the `dim()` function:

```{r col-row-len}

nrow(stations)
ncol(stations)
dim(stations)

```



Whenever we see rectangular data like this in R, it's probably a `data.frame` object, but just to check, we can always ask R to tell us what the class of the object is:

```{r check-class}
class(stations)
```

The printed output shows us the first few rows of the stations data we just read, but if we wanted to dig a bit deeper and see more than 10 rows and 7 columns of data, we can use the function `View()`, which in RStudio opens a data viewer. 

```{r view-data, echo=TRUE, eval=FALSE}

View(stations)

```

Within the viewer, we can search the data.frame, sort rows, and scroll through the data to inspect it.

A data.frame is made of many vectors of the same length. We can access a column using the `$` operator, and subset the vector with bracket notation `[`. To access the first row in the `WELL_TYPE` column: 

```{r dollar-sign}

stations$WELL_TYPE[1]

```

If we wanted the `WELL_TYPE` entries 1 through 10, we can subset by a vector of the sequence from 1 through 10:

```{r dollar-sign-subset}

stations$WELL_TYPE[1:10]

```

Sometimes it's helpful to count unique variables in a column, especially for categorical data such as the well type. 

```{r table}
table(stations$WELL_TYPE)
```


<br>  

### xlsx and xls (Excel files)

Excel files are very common, and R has great utilities for reading in and processing excel files. [Calenviroscreen data](https://oehha.ca.gov/calenviroscreen) comes in excel format, which we can read in like so:

```{r import-excel, warning=FALSE}
library(readxl)
ces <- read_xlsx(here("data", "calenviroscreen", "ces3results.xlsx"))
head(ces, 10) # print the first 10 rows
```

By default, `read_xlsx()` reads in the first sheet. However, there may be many sheets in an excel file. If we want to read in a different sheet, we can tell R which sheet to read in, and even how many lines to skip before reading in data. 

```{r import-excel-2}
metadata <- read_xlsx(here("data", "calenviroscreen", "ces3results.xlsx"), 
                      sheet = 2, 
                      skip  = 6)
metadata
```



:::challenge

<font color="#009E73">**Challenge 1**</font> 

1. open the documentation for `read_xlsx()` using `?` (Hint: type `?read_xlsx` in the console and hit enter)
2. read through the Arguments to get a sense of what else the function can do
3. Read in the "Index Summary" sheet of `here("data", "healthy_watersheds", "CA_PHWA_TabularResults_170518.xlsx")`, and select the appropriate number of rows to skip. 
4. What happened to column names during the read? 

:::

<br>

<details>
  <summary class="challenge-ans-title"><font color="#0072B2">**Click for Answers!**</font></summary>
  <div class="challenge-ans-body">

```{r import-excel-challenge-1}
health <- read_xlsx(here("data", "healthy_watersheds", "CA_PHWA_TabularResults_170518.xlsx"), 
                    sheet = 2, 
                    skip  = 4)
head(health)
```

Some column names during the read were renamed because they're the same in the Excel sheet. In R, a data.frame can have only one unique name per column -- duplicate names aren't allowed! Thus, R renamed those duplicate names. In a later module, we will see how to rename columns within R.

  </div>
</details>


<br>

### shp (ArcGIS shapefiles)

Geospatial data is ubiquitous. So is the ArcGIS data format, the shapefile. A georeferenced shapefile is, at minimum made of 4 files: `.shp`, `.prj`, `.dbf`, and `.shx`. 

```{r, read-shp, echo=TRUE, eval=TRUE}

# unzip Sacramento county shapefile
unzip(here("data", "shp", "sac_county.zip"), exdir = here("data", "shp", "sac"))

# read the shapefile
library(sf)
sac_county <- st_read(here("data", "shp", "sac", "sac_county.shp"))

```


```{r plot-sac}

library(ggplot2)
ggplot(sac_county) + geom_sf()

```


### dbf

`.dbf` files are one kind of database file. If you've ever opened a shapefile with attribute information, you've used a `.dbf` file. The `foreign` package allows us to read `.dbf` files into R. Since this is a new package, we need to install it with `install.packages("foreign")`. 

We've been loading entire packages with the `library()` function, but you can also call a function from a package without loading it by using `<package_name>::<function_name>` syntax. Let's load the `.dbf` file from our Sacramento County polygon.  

```{r import-dbf}
foreign::read.dbf(here("data", "shp", "sac", "sac_county.dbf"))
```


### rds and .rda

`.rds` and `.rda` (`.rda` is shorthand for `.RData`) are a special R-based data formats used to store R objects. These files can be read just like another other import functions shown above. Let's use it to import the groundwater level station data we read in earlier. Note that a `.rds` file can hold any single R object.

```{r, read-rds}
stations <- read_rds(here("data","gwl","stations.rds"))
head(stations)
```

Sometimes, you may create an intermediate result that is time-consuming to recreate from scratch each time, and you want to save that intermediate result to streamline future analyses. You can export, or write this object to any number of data formats like a csv, SQL database, or shapefile. Unlike these data formats however, data saved as .rds are saved as one of R's object classes, like data.frame, vector, list, and so on. In practice, only R is used to read .rds and .rda files, so these formats are chosen when we expect to use R to read these data at a later time.


### sqlite

SQLite is an open-source database format based on SQL that's useful for storing large datasets locally on your computer. The methods to connect to a SQLite database, list tables, read tables, and send queries are similar across other cloud databases you may encounter in the wild, like [Postgres](https://rpostgres.r-dbi.org/), and enterprise database systems like Microsoft Access.

```{r, read-sql}
library(RSQLite)

# location of an sqlite database
dbpath <- here("data", "gwl", "gwl_data.sqlite")

# actually connect to the database
dbcon <- dbConnect(dbDriver("SQLite"), dbpath)

# list all the tables in the database
dbListTables(dbcon)

# get one of the tables into a dataframe
head(dbReadTable(dbcon, "stations"))
head(dbReadTable(dbcon, "measurements_sep"))

# pass a query to the database
dbGetQuery(dbcon, "SELECT * from measurements_sep WHERE STN_ID = 4775 LIMIT 5")
```


## Write (export) data

To write (export) data in R you need 2 things: data to write, and a location and format to write the data. For example, if we wanted to write our stations data to a csv in the "results" folder, we would do the following:

```{r, write-csv, eval = FALSE}
# write "stations" to a file in the results folder called "my_stations.csv"
write_csv(stations, here("results", "my_stations.csv"))
```

Now check that location and verify that your station data was written. 

We can do the same for other files:

```{r, write-other, eval = FALSE}
# write the Sacramento county polygon to a shapefile
st_write(sac_county, here("results", "sac_county.shp"))

# write the Sacramento county polygon to an rds file
write_rds(sac_county, here("results", "sac_county.rds"))
```

As before, navigate to these folders to verify these data were written. We can also check to see if these data exist from within R:

```{r verify-write-other, eval = FALSE}
my_results <- list.files(here("results"))
my_files  <- c("sac_county.shp", "sac_county.rds")
 
# test if your files are in the results folder
my_files %in% my_results

# another handy function is `file.exists`, which tells you if your file exists
file.exists(here("results", "sacramento_county.shp"))
file.exists(here("results", "sac_county.shp"))
file.exists(here("results", "sac_county.rds"))
```



:::challenge

<font color="#009E73">**Challenge 2**</font> 

1. create an object called `breakfast` and assign it a string with what you had for breakfast. 
2. write that object to `breakfast.rds` file in `/results`
3. verify this worked by reading the string back in and inspecting it

:::

<br>

<details>
  <summary class="challenge-ans-title"><font color="#0072B2">**Click for Answers!**</font></summary>
  <div class="challenge-ans-body">

```{r challenge-2, eval=FALSE}

# create a string and write it to an rds file
breakfast <- "green eggs and ham"

# write_rds takes two arguments: the object to write and location to wite it
write_rds(breakfast, here("results", "breakfast.rds"))

# read the rds file back into R and save it as a variable
my_breakfast <- read_rds(here("results", "breakfast.rds"))

# use the `cat()` function (concatenate) to announce your breakfast
cat("Today for breakfast I ate", my_breakfast)

```

  </div>
</details>

<br>  

<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/4.0.0/css/bootstrap.min.css" integrity="sha384-Gn5384xqQ1aoWXA+058RXPxPg6fy4IWvTNh0E263XmFcJlSAwiGgFAW/dAiS6JXm" crossorigin="anonymous">

<a href="m_project_management.html" class="btn btn-secondary" style="float: left">Previous module:<br> 3. Project management</a>
<a href="m_ggplot.html" class="btn btn-secondary" style="float: right;">Next module:<br> 5. Data visualization</a>  

