---
title: "Interactive Visualization"
description: | 
   Having fun with interactivity in R
output: 
  distill::distill_article:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}

library(knitr)
library(glue)
library(here)

```

::: {.obj}
**Learning objectives**

-  How to use interactive visualization in R
-  Workflows to go from data to visualization
-  Understand a visualization objective
:::


# Why interactive visualizations?

One of the most exciting and rewarding parts of learning to code is the ability to ***see results***! Visualization is one of `R`'s strengths, and there are a seemingly constantly growing number of packages that can assist you in customizing just about any visualization you can think of. Importantly, this allows users to access and interact with data on a level that would not otherwise be possible.

We can learn new things and develop new questions by effectively interacting with data...so the ability to stitch raw data into something visual can help provide unique insight that may be much harder to discern with a table or some raw data.

Let's demonstrate a simple way we can collect data, update it *automagically*, and visualize in a fairly streamlined manner. Keep in mind, this is one approach, but there are many, with increasing complexity. But when starting out, working examples (as in, it works when you try it!) are the best way to build your proficiency in R programming.

## Mapping with Geocoder

This example will couple the power of collecting user input (favorite places to eat/drink) with interactive mapping. We'll use a Form to collect some data, and then display that via a map. This illustrates some of the power (and fun!) of programming in R.

### Collect Data

Answer as many questions as you feel comfortable, but at minimum, the street address/name of your favorite place to eat or get a beverage (hot or cold).

<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSfy6hsEVZ2b0yvAnK907ESR83j0M8jqEM7cFyXKjKpo8GGwnw/viewform?embedded=true" width="640" height="1026" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>


### Import and Map Data

We'll need these packages to work with this data.

```{r pkgs, message=FALSE, warning=FALSE}

library(tidygeocoder) # geocode our addresses
library(tidyverse) # wrangle data
library(janitor) # clean column names
library(sf) # make spatial data
library(mapview) # interactive maps!
mapviewOptions(fgb = FALSE)


```

Here's the whole script we need! To generate a public `.csv` from a Form, we can use `File > Publish to the Web`. Importantly, here we are publishing to a `.csv` so we can read the data in directly and simply using `read_csv`. We then clean, make it into an `sf` object, and map!

```{r mapview, layout="1-page", echo=TRUE, eval=TRUE}

# the url for the data:
form_data <- paste0("https://docs.google.com/spreadsheets/d/e/",
                    "2PACX-1vSODxBm_z5Gu8a42C6ZFEa3S5iTbYV-",
                    "qucCGvasGS6c0qFUAml5vSMEgbvI9PYo1HJ20Y_WY62aTAb-",
                    "/pub?gid=1462593645&single=true&output=csv")

# read in url and clean
dat <- read_csv(form_data) %>% clean_names() %>% rename(address=3)

# geocode
dat_geo <- dat %>%
  geocode(address, method = 'osm', lat = latitude , long = longitude)

# make into sf object so we can map
dat_geo <- dat_geo %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs=4269, remove = FALSE )

# map!
mapview(dat_geo, zcol="comfort_using_r", layer.name="R comfort level", cex=6.5)

```


# Other Examples


Introduce data
Mapview: easy to use to plot a first cut of spatial data, great for EDA
Leaflet: more advanced visualization. Markers, popups, labels
Mention how to put this into a dashboard or Rmd html_document
Flexdashboard
 


