---
title: "Interactive Visualization"
description: | 
   Having fun with interactivity in R
output: 
  distill::distill_article:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}

library(knitr)
library(glue)
library(here)

```

::: {.obj}
**Learning objectives**

-  How to use interactive visualization in R
-  Workflows to go from data to visualization
-  Understand a visualization objective
:::


# Why interactive visualizations?

One of the most exciting and rewarding parts of learning to code is the ability to ***see results***! Visualization is one of `R`'s strengths, and there are a seemingly constantly growing number of packages that can assist you in customizing just about any visualization you can think of. Importantly, this allows users to access and interact with data on a level that would not otherwise be possible.

We can learn new things and develop new questions by effectively interacting with data...so the ability to stitch raw data into something visual can help provide unique insight that may be much harder to discern with a table or some raw data.

Let's demonstrate a simple way we can collect data, update it *automagically*, and visualize in a fairly streamlined manner. Keep in mind, this is one approach, but there are many, with increasing complexity. But when starting out, working examples (as in, it works when you try it!) are the best way to build your proficiency in R programming.

## Mapping with Geocoder

This example will couple the power of collecting user input (favorite places to eat/drink) with interactive mapping. We'll use a Form to collect some data, and then display that via a map. This illustrates some of the power (and fun!) of programming in R.

### Collect Data

Answer as many questions as you feel comfortable, but at minimum, the street address/name of your favorite place to eat or get a beverage (hot or cold).

<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSfy6hsEVZ2b0yvAnK907ESR83j0M8jqEM7cFyXKjKpo8GGwnw/viewform?embedded=true" width="640" height="1026" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>


### Import and Map Data

We'll need these packages to work with this data.

```{r pkgs, message=FALSE, warning=FALSE}

library(tidygeocoder) # geocode our addresses
library(tidyverse) # wrangle data
library(janitor) # clean column names
library(glue)
library(sf) # make spatial data
library(mapview) # interactive maps!
mapviewOptions(fgb = FALSE)


```

Here's the whole script we need! To generate a public `.csv` from a Form, we can use `File > Publish to the Web`. Importantly, here we are publishing to a `.csv` so we can read the data in directly and simply using `read_csv`. We then clean, make it into an `sf` object, and map!

```{r mapview, layout="1-page", echo=TRUE, eval=TRUE}

# the url for the data:
form_data <- paste0("https://docs.google.com/spreadsheets/d/e/",
                    "2PACX-1vSODxBm_z5Gu8a42C6ZFEa3S5iTbYV-",
                    "qucCGvasGS6c0qFUAml5vSMEgbvI9PYo1HJ20Y_WY62aTAb-",
                    "/pub?gid=1462593645&single=true&output=csv")

# read in url and clean
dat <- read_csv(form_data) %>% clean_names() %>% rename(address=3)

# geocode
dat_geo <- dat %>%
  geocode(address, method = 'osm', lat = latitude , long = longitude)

# make into sf object so we can map
dat_geo <- dat_geo %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs=4326, remove = FALSE )

# map!
mapview(dat_geo, zcol="comfort_using_r", layer.name="R comfort level", cex=6.5)

```

### Use `{leaflet}`

We used {`mapview`} above, but we could also use the `{leaflet}` package to make a more customizable map with some fancy icons, and a measuring tool, to see how far each location is from wherever it is you are interested. Or what the the total area is that encompasses all our points.

```{r leaflet, layout="1-page"}

library(leafpm)
library(leaflet)
library(leaflet.extras)
library(htmltools)

# set up our map
m <- leaflet() %>%
  # add tiles or the "basemaps"
  addTiles(group = "OSM") %>% # defaults to Open Street Maps
  addProviderTiles(providers$CartoDB.Positron, group="Positron") %>% 
  addProviderTiles(providers$Stamen.TonerLite, group = "Toner Lite") %>%
  addMarkers(icon=icons(
    iconUrl = "https://leafletjs.com/examples/custom-icons/leaf-green.png",
    shadowUrl = "https://leafletjs.com/examples/custom-icons/leaf-shadow.png",
    iconWidth = 38, iconHeight = 95, shadowWidth = 50, shadowHeight = 64,
    iconAnchorX = 22, iconAnchorY = 94, shadowAnchorX = 4, shadowAnchorY = 62,
    popupAnchorX = -3, popupAnchorY = -76),
  lng=-121.4944, lat=38.5816, 
  popup="Sacramento!", group = "Home") %>% 
  addMarkers(data = dat_geo, group = "Food & Drink",
             label = ~htmlEscape(first_name),
             popup = glue("<b>Name:</b> {dat_geo$first_name}<br>
                          <b>Address:</b> {dat_geo$address}<br>
                          <b>R comfort (1-10):</b>{dat_geo$comfort_using_r}")
             )  %>% 
  addLayersControl(
    baseGroups = c("Toner Lite", "Positron", "OSM"),
    overlayGroups = c("Home", "Food & Drink"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  addMeasure()

m  # Print the map

```


## Using D3

A powerful visualization tool is the [D3.js](https://d3js.org) (javascript) library, which has some amazing options. What's great is there are ways to easily translate your R code into D3, via packages like [{`r2d3`}](https://rstudio.github.io/r2d3/index.html).

Here's an example of a calendar visualization, based on the stock market open and close between 2006 and 2010.

```{r, eval=TRUE, echo=FALSE}
library(r2d3)

r2d3(data = read_csv("https://raw.githubusercontent.com/rstudio/r2d3/master/vignettes/gallery/calendar/dji-latest.csv"), d3_version = 4, container = "div", options = list(start = 2006, end = 2011), script = here("functions","calendar.js"))
```

### A Network with D3

Another example of a way to interactively engage with data is networks. Here's a simple example of an interactive network visualization:

```{r d3, layout="1-page"}
# Libraries
library(igraph)
library(networkD3)

# create a dataset:
data <- tibble(
  from=c("Dam","Dam","Dam", "Dam",
         "River","River","River", "River","River",
         "Canal", "Canal", 
         "Diversion","Diversion", 
         "Reservoir", "Reservoir","Reservoir",
         "Lake","Lake","Lake", "Lake", 
         "Road","Road","Road",
         "Culvert", "Culvert",
         "Fish", "Fish","Fish",
         "Frog","Frog","Frog",
         "MacroInvertebrates","MacroInvertebrates"),
  to=c("River","Reservoir","Canal","Diversion",
       "Lake","Reservoir","Frog","Fish","MacroInvertebrates",
       "Diversion", "Reservoir",
       "Dam", "River",
       "River","Dam","Fish",
       "Fish","Dam","Frog","MacroInvertebrates",
       "Fish","Dam", "Canal",
       "Road", "Dam",
       "Frog", "River","MacroInvertebrates",
       "Fish", "River", "Lake",
       "River", "Lake")
)

# Plot
(p <- simpleNetwork(data, height="600px", width="600px", 
                    fontSize = 16, fontFamily = "serif",
                    nodeColour = "darkblue", linkColour = "steelblue",
                    opacity = 0.9, zoom = TRUE, charge=-500))

```

If we wanted to save this interactive visualization and send it as a standalone `.html` file, we could use the following code:

```{r, eval=FALSE}

# save the widget
library(htmlwidgets)
saveWidget(p, file=here("output","network_interactive.html"))

```


## Plotly

Another great tool for interacting with data is the `{plotly}` package, which if using the `{ggplot2}` framework, has a very handy function you can wrap around pretty much any `ggplot`: `ggplotly()`. This allows you to click, hover, zoom, etc. inside your plot, and is a very useful tool for exploring data.

Here's a quick example, say you were asked to find the date of a peak flow for a given year, or an outlier in a graph? We could figure this out with some `dplyr::filter` and additional queries of our data, but it would also be easy to visually look and see what this is on a plot. This is where the power of `plotly` can be very helpful.

Let's use the [CalEnviroScreen data](https://oehha.ca.gov/calenviroscreen/report/calenviroscreen-30) from Sacramento County from a [previous module](). Recall, higher scores mean a higher impact of pollution to a community. 

```{r ciplot, layout="1-page", eval=TRUE, echo=TRUE, fig.cap="Static plot of CES data...how do we interact with these data?"}

# load CES data for Sacramento county
ces3_sac <- readRDS(here::here("data", "ces3_sac.rds"))

mapview(ces3_sac, zcol="CIscoreP")

```

Let's make a static plot that shows the association between groundwater threats and the CES score. The trend is interesting, but what if we want to know what census tracts are associated with outliers (high or low) on this static plot?

```{r, layout="1-page"}
# plot of Groundwater threats vs. CES score
(ces_plot <- ggplot(data = ces3_sac, 
                    aes(x = gwthreatsP, y = CIscoreP, label=tract)) + 
    geom_point() +
    geom_smooth(method = "gam") +
    cowplot::theme_half_open() +
    labs(title="CalEnviroscreen Score vs. 
         Groundwater Threats in Sacramento County",
         subtitle="Higher CI Score indicates higher threat or impact to community",
         x="Groundwater Threats (percentile)", y="CI Score (percentile)"))
  
```

Try wrapping the exact plot above with `ggplotly()`. Now we can quickly interact and select data of interest. Try hovering your pointer over a point on the plot below. Or, click and drag to select a specific region of the plot to zoom into just that region. You can double click to return to the original zoom level.

```{r plotlyCI, layout="1-page", eval=TRUE, echo=TRUE}

library(plotly)
ggplotly(ces_plot)


```


## Summary

There are many options to make very fancy visualizations in R, but remember to consider what the objective is, and how you can frame things for the end user. We like stories, so while a really cool visualization may definitely grab someone's attention, it can be much more powerful if it's coupled with a story. Or, sometimes data visualization can be a really important tool for Exploratory Data Analysis (see [EDA](../intro/m_exploratory_DA.html)). 

Whatever the visualization, if it communicates data in a way that provides new insight, or provides a easier medium to transfer information (i.e., with an Rmarkdown document), the ability to quickly take data, transform it, and visualize it is a powerful skill to have.



## More Resources

 - [Rmarkdown Flexdashboard](https://pkgs.rstudio.com/flexdashboard/)
 - [R2D3 Visualizations](https://rstudio.github.io/r2d3/index.html)
 - [RGraph Gallery](https://www.r-graph-gallery.com/index.html)
 - All things [Shiny](https://shiny.rstudio.com/gallery/)
 - [Network visualization](https://kateto.net/network-visualization)



<br> 

<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/4.0.0/css/bootstrap.min.css" integrity="sha384-Gn5384xqQ1aoWXA+058RXPxPg6fy4IWvTNh0E263XmFcJlSAwiGgFAW/dAiS6JXm" crossorigin="anonymous">

<a href="index.html" class="btn btn-secondary" style="float: left">Previous module:<br>Version Control</a> <a href="m_version_control.html" class="btn btn-secondary" style="float: right;">Next module:<br>Reproducible Workflows</a>




